---
title: "Assessing the spatial patterning of Middle Paleolithic human settlement in Western Iberia"
author:
  - João Cascalheira
  - Célia Gonçalves
  - Daniela Maio
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  Middle Paleolithic; Settlement patterns; GIS; Western Iberia
---


<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

```{r, setup, eval = FALSE, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

#library(MidPalSet) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` directory

library(rgdal)
library(raster)
library(rasterVis)
library(lattice)
library(foreach)
library(rnaturalearth)
library(tidyverse)
library(ggplot2)
library(broom)
library(foreach)
library(sf)
library(osfr)
library(cowplot)
library(gridExtra)
library(rela)
library(psych)
library(factoextra)

```


```{r load-data, eval = FALSE, echo = FALSE}

#Download geo tiffs from OSF repositorium when running this document for the first time

#osf_retrieve_file("nv6r8") %>%
#  osf_download(path = "../data/raw_data")
#osf_retrieve_file("jb6v8") %>%
#  osf_download(path = "../data/raw_data")
#osf_retrieve_file("yxk23") %>%
#  osf_download(path = "../data/raw_data")
#osf_retrieve_file("57qfn") %>%
#  osf_download(path = "../data/raw_data")
#osf_retrieve_file("mx2bt") %>%
#  osf_download(path = "../data/raw_data")

#Import geo tiffs
areaDEM <- raster("../data/raw_data/areaDEM.tif")
areaDEM_cut <- raster("../data/raw_data/areaDEM_cut.tif")
areaSLOPE_cut <- raster("../data/raw_data/areaSLOPE_cut.tif")
areaASPECT_cut <- raster("../data/raw_data/areaASPECT_cut.tif")
areaDIST_cut <- raster("../data/raw_data/areaDIST_cut2.tif")


# Convert coordinates to longlat WGS84
areaDEM <- projectRaster(areaDEM, crs="+proj=longlat +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
areaDEM_cut <- projectRaster(areaDEM_cut, crs="+proj=longlat +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
areaSLOPE_cut <- projectRaster(areaSLOPE_cut, crs="+proj=longlat +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
areaASPECT_cut <- projectRaster(areaASPECT_cut, crs="+proj=longlat +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
areaDIST_cut <- projectRaster(areaDIST_cut, crs="+proj=longlat +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")


#Import sites and convert to shapefile
#sites <- read.csv("../data/raw_data/sites_short_list.csv", header=TRUE)
#caves <- read.csv("../data/raw_data/all_caves.csv", header=TRUE)

#Convert to shapefile
#coordinates(sites)<-~Longitude+Latitude
#proj4string(sites) = CRS("+init=epsg:4326")
#writeOGR(sites, "../data/raw_data/spatial_data/shapefiles","sites_short", "ESRI Shapefile")

#coordinates(caves)<-~Longitude+Latitude
#proj4string(caves) = CRS("+init=epsg:4326")
#writeOGR(caves, "../data/raw_data/spatial_data/shapefiles","caves", "ESRI Shapefile")


#Read shapefiles
sites_short <- readOGR(dsn="../data/raw_data/spatial_data/shapefiles", layer="sites_short")
caves <- readOGR(dsn="../data/raw_data/spatial_data/shapefiles", layer="caves")
#rivers <- readOGR(dsn="../data/raw_data/spatial_data/shapefiles", layer="RHidro_Rivers_v2")

#Filter sites by Type
sites_sub <- sites_short[sites_short$Type == "Cave" | sites_short$Type == "Open air",]
sites_sub$Type <- factor(sites_sub$Type)

#Convert coordinates to longlat WGS84
sites_sub_long <- spTransform(sites_sub, "+proj=longlat +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")

caves_long <- spTransform(caves, "+proj=longlat +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")

```



```{r extract-variables, eval = FALSE, echo = FALSE}

# Extract variables

# Resample Distance raster
areaDIST_cut <- resample(areaDIST_cut,areaDEM_cut, resample='bilinear')

# Stack maps
terrainstack <- stack(areaDEM_cut,
                      areaSLOPE_cut,
                      areaASPECT_cut,
                      areaDIST_cut)

# Extract mean values from stack for each site and all caves within a 20 m radius
sites_vals <- raster::extract(terrainstack,
                      sites_sub_long,
                      buffer = 20,
                      fun = mean,
                      sp = TRUE)

caves_vals <- raster::extract(terrainstack,
                              caves_long,
                              buffer = 20,
                              fun = mean,
                              sp = TRUE)


# Subdivide sites for further analysis
open_air <- dplyr::filter(as.data.frame(sites_vals), Type == "Open air")
caves <- dplyr::filter(as.data.frame(sites_vals), Type == "Cave")
all_caves <- as.data.frame(caves_vals)

```


# Introduction


```{r general-map, eval = FALSE, echo = FALSE}

rasterVis::levelplot(areaDEM,
                     margin = list(x = FALSE,
                                   y = TRUE),
                     col.regions = terrain.colors(16),
                     xlab = list(label = "",
                                 vjust = -0.25),
                     sub = list(
                       label = "Meters a.s.l.",
                       font = 1,
                       cex = .9,
                       hjust = 0.5),
                     colorkey=list(space="bottom"),
                     key = list(
                       space = "left",
                       points = list(
                         pch = c(18,20),
                         col = c("red","blue")),
                       text = list(
                         c("Cave","Open-air"),
                         cex=1))) +
  spplot(sites_sub_long, # add a layer of points
         zcol = "Type",
         cex = 2,
         pch = c(18,20),
         col.regions = c("red","blue"))

```


# Methods

## Site selection

## Mapping and GIS analysis

## Statistical analysis


# Results

## Open air

```{r plot-altitude, eval = FALSE, echo = FALSE}

# Calculate site altitude densities
open_air_altitude_densities <- open_air %$%
   areaDEM_cut %>%
   density(from = -26, to = 1993, n = 1965) %>%
   broom::tidy() %>%
   tibble::as_tibble() %>%
   dplyr::mutate(y = y * 1965) %>%
   dplyr::rename(Elevation = x,
                Frequency = y)

# Load background values into memory for fast resampling
background_altitude_values <- areaDEM_cut %>%
  values() %>%
  na.omit()

# Draw 1000 random samples from background and calculate their densities
background_altitude_densities <- foreach::foreach(n = 1:1000, .combine = rbind) %do% {
  background_altitude_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    density(from = -26, to = 1993, n = 1965) %>%
    broom::tidy() %>%
    tibble::as_tibble() %>%
    dplyr::mutate(y = y * 1965)
} %>%
  dplyr::group_by(x) %>%
  purrrlyr::by_slice(function(x){
    quantile(x$y, probs = c(0.025, 0.5, 0.975)) %>%
      t() %>%
      broom::tidy()
  }, .collate = "cols") %>%
  magrittr::set_names(c("Elevation", "Lower CI", "Frequency", "Upper CI"))


# Plot distributions
open_air_elevation <- ggplot() +
  geom_line(data = background_altitude_densities,
            mapping = aes(x = Elevation,
                          y = Frequency)) +
  geom_ribbon(data = background_altitude_densities,
              mapping = aes(x = Elevation,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_altitude_densities,
            mapping = aes(x = Elevation,
                          y = Frequency),
            color = "red") +
  theme_cowplot() +
  ylab("") + xlab("meters")


```

```{r test-altitude, echo = false, eval = FALSE}

# Draw 1000 random samples from background, and compute two-sample Wilcoxon tests (Mann-Whitney U tests)
background_altitude_MWU <- foreach(n = 1:1000, .combine = rbind) %do% {
  background_sample <- background_altitude_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    wilcox.test(x = open_air$areaDEM_cut,
                y = .,
                exact = FALSE) %>%
    broom::tidy() %>%
    tibble::as_tibble()

} %>%
  dplyr::select(statistic, p.value)


# Get the median test statistic and 95% confidence interval
background_altitude_MWU <- foreach::foreach(prob = c(0.025,0.5,0.975), .combine = rbind) %do% {
  background_altitude_MWU %>%
    dplyr::summarise_all(quantile, probs = prob)
} %>%
  t() %>%
  magrittr::set_colnames(c("Lower CI","Median","Upper CI")) %>%
  magrittr::set_rownames(c("U statistic","p-value"))


round(background_altitude_MWU, 3)

```


```{r plot-slope, eval = FALSE, echo = FALSE}

# Calculate site altitude densities
open_air_slope_densities <- open_air %$%
   areaSLOPE_cut %>%
   density(from = 0, to = 70, n = 35) %>%
   broom::tidy() %>%
   tibble::as_tibble() %>%
   dplyr::mutate(y = y * 35) %>%
   dplyr::rename(Slope = x,
                Frequency = y)

# Load background values into memory for fast resampling
background_slope_values <- areaSLOPE_cut %>%
  values() %>%
  na.omit()

# Draw 1000 random samples from background and calculate their densities
background_slope_densities <- foreach::foreach(n = 1:1000, .combine = rbind) %do% {
  background_slope_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    density(from = 0, to = 70, n = 35) %>%
    broom::tidy() %>%
    tibble::as_tibble() %>%
    dplyr::mutate(y = y * 35)
} %>%
  dplyr::group_by(x) %>%
  purrrlyr::by_slice(function(x){
    quantile(x$y, probs = c(0.025, 0.5, 0.975)) %>%
      t() %>%
      broom::tidy()
  }, .collate = "cols") %>%
  magrittr::set_names(c("Slope", "Lower CI", "Frequency", "Upper CI"))


# Plot distributions
open_air_slope <- ggplot() +
  geom_line(data = background_slope_densities,
            mapping = aes(x = Slope,
                          y = Frequency)) +
  geom_ribbon(data = background_slope_densities,
              mapping = aes(x = Slope,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_slope_densities,
            mapping = aes(x = Slope,
                          y = Frequency),
            color = "red") +
  theme_cowplot() +
  ylab("") + xlab("degrees")


```

```{r test-slope, echo = false, eval = FALSE}

# Draw 1000 random samples from background, and compute two-sample Wilcoxon tests (Mann-Whitney U tests)
background_slope_MWU <- foreach(n = 1:1000, .combine = rbind) %do% {
  background_sample <- background_slope_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    wilcox.test(x = open_air$areaSLOPE_cut,
                y = .,
                exact = FALSE) %>%
    broom::tidy() %>%
    tibble::as_tibble()

} %>%
  dplyr::select(statistic, p.value)


# Get the median test statistic and 95% confidence interval
background_slope_MWU <- foreach::foreach(prob = c(0.025,0.5,0.975), .combine = rbind) %do% {
  background_slope_MWU %>%
    dplyr::summarise_all(quantile, probs = prob)
} %>%
  t() %>%
  magrittr::set_colnames(c("Lower CI","Median","Upper CI")) %>%
  magrittr::set_rownames(c("U statistic","p-value"))


round(background_slope_MWU, 3)

```


```{r plot-aspect, eval = FALSE, echo = FALSE}

# Calculate site altitude densities
open_air_aspect_densities <- open_air %$%
   areaASPECT_cut %>%
   density(from = -1, to = 360, n = 181) %>%
   broom::tidy() %>%
   tibble::as_tibble() %>%
   dplyr::mutate(y = y * 181) %>%
   dplyr::rename(Aspect = x,
                Frequency = y)

# Load background values into memory for fast resampling
background_aspect_values <- areaASPECT_cut %>%
  values() %>%
  na.omit()

# Draw 1000 random samples from background and calculate their densities
background_aspect_densities <- foreach::foreach(n = 1:1000, .combine = rbind) %do% {
  background_aspect_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    density(from = -1, to = 360, n = 181) %>%
    broom::tidy() %>%
    tibble::as_tibble() %>%
    dplyr::mutate(y = y * 181)
} %>%
  dplyr::group_by(x) %>%
  purrrlyr::by_slice(function(x){
    quantile(x$y, probs = c(0.025, 0.5, 0.975)) %>%
      t() %>%
      broom::tidy()
  }, .collate = "cols") %>%
  magrittr::set_names(c("Aspect", "Lower CI", "Frequency", "Upper CI"))


# Plot distributions
open_air_aspect <- ggplot() +
  geom_line(data = background_aspect_densities,
            mapping = aes(x = Aspect,
                          y = Frequency)) +
  geom_ribbon(data = background_aspect_densities,
              mapping = aes(x = Aspect,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_aspect_densities,
            mapping = aes(x = Aspect,
                          y = Frequency),
            color = "red") +
  theme_cowplot() +
  ylab("") + xlab("degrees")


```

```{r test-aspect, echo = false, eval = FALSE}

# Draw 1000 random samples from background, and compute two-sample Wilcoxon tests (Mann-Whitney U tests)
background_aspect_MWU <- foreach(n = 1:1000, .combine = rbind) %do% {
  background_sample <- background_aspect_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    wilcox.test(x = open_air$areaASPECT_cut,
                y = .,
                exact = FALSE) %>%
    broom::tidy() %>%
    tibble::as_tibble()

} %>%
  dplyr::select(statistic, p.value)


# Get the median test statistic and 95% confidence interval
background_aspect_MWU <- foreach::foreach(prob = c(0.025,0.5,0.975), .combine = rbind) %do% {
  background_aspect_MWU %>%
    dplyr::summarise_all(quantile, probs = prob)
} %>%
  t() %>%
  magrittr::set_colnames(c("Lower CI","Median","Upper CI")) %>%
  magrittr::set_rownames(c("U statistic","p-value"))


round(background_aspect_MWU, 3)

```


```{r plot-distance, eval = FALSE, echo = FALSE}

# Calculate site altitude densities
open_air_distance_densities <- open_air %$%
   areaDIST_cut %>%
   density(from = 0, to = 7697, n = 350) %>%
   broom::tidy() %>%
   tibble::as_tibble() %>%
   dplyr::mutate(y = y * 350) %>%
   dplyr::rename(Distance = x,
                Frequency = y)

# Load background values into memory for fast resampling
background_distance_values <- areaDIST_cut %>%
  values() %>%
  na.omit()

# Draw 1000 random samples from background and calculate their densities
background_distance_densities <- foreach::foreach(n = 1:1000, .combine = rbind) %do% {
  background_distance_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    density(from = 0, to = 7697, n = 350) %>%
    broom::tidy() %>%
    tibble::as_tibble() %>%
    dplyr::mutate(y = y * 350)
} %>%
  dplyr::group_by(x) %>%
  purrrlyr::by_slice(function(x){
    quantile(x$y, probs = c(0.025, 0.5, 0.975)) %>%
      t() %>%
      broom::tidy()
  }, .collate = "cols") %>%
  magrittr::set_names(c("Distance", "Lower CI", "Frequency", "Upper CI"))


# Plot distributions
open_air_distance <- ggplot() +
  geom_line(data = background_distance_densities,
            mapping = aes(x = Distance,
                          y = Frequency)) +
  geom_ribbon(data = background_distance_densities,
              mapping = aes(x = Distance,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_distance_densities,
            mapping = aes(x = Distance,
                          y = Frequency),
            color = "red") +
  theme_cowplot() +
  ylab("") + xlab("meters")


```

```{r test-aspect, echo = false, eval = FALSE}

# Draw 1000 random samples from background, and compute two-sample Wilcoxon tests (Mann-Whitney U tests)
background_distance_MWU <- foreach(n = 1:1000, .combine = rbind) %do% {
  background_sample <- background_distance_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    wilcox.test(x = open_air$areaDIST_cut,
                y = .,
                exact = FALSE) %>%
    broom::tidy() %>%
    tibble::as_tibble()

} %>%
  dplyr::select(statistic, p.value)


# Get the median test statistic and 95% confidence interval
background_distance_MWU <- foreach::foreach(prob = c(0.025,0.5,0.975), .combine = rbind) %do% {
  background_distance_MWU %>%
  dplyr::summarise_all(quantile, probs = prob)
} %>%
  t() %>%
  magrittr::set_colnames(c("Lower CI","Median","Upper CI")) %>%
  magrittr::set_rownames(c("U statistic","p-value"))


round(background_distance_MWU, 3)

```

```{r open-air-plots, eval = FALSE, echo = FALSE}

plot_grid(open_air_elevation, open_air_slope, open_air_aspect, open_air_distance, labels = c("A", "B", "C", "D"), nrow = 2, ncol = 2)

```



## Caves


```{r caves-boxplots, eval = FALSE, echo = FALSE}


cave <- caves %>% 
  dplyr::select(SITE, areaDEM_cut, areaSLOPE_cut, areaASPECT_cut, areaDIST_cut2) %>% 
  mutate(TYPE = "MP Caves")


all_caves <- all_caves %>% 
  rename(SITE = Ã.__SITE) %>% 
  dplyr::select(SITE, areaDEM_cut, areaSLOPE_cut, areaASPECT_cut, areaDIST_cut2) %>% 
  mutate(TYPE = "All Caves")

cave <- rbind(cave, all_caves)

caves_elevation <- cave %>%
  mutate(TYPE = fct_relevel(TYPE, 
            "MP Caves", "All Caves")) %>% 
  ggplot(aes(x = TYPE, y = areaDEM_cut, fill = TYPE)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter() +
  ylab("meters") + xlab("") +
  theme_cowplot() +
  theme(legend.position="none") +
  scale_fill_brewer(palette = "Dark2")


caves_slope <- cave %>%
  mutate(TYPE = fct_relevel(TYPE, 
            "MP Caves", "All Caves")) %>%
  ggplot(aes(x = TYPE, y = areaSLOPE_cut, fill = TYPE)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter() +
  ylab("degrees") + xlab("") +
  theme_cowplot() +
  theme(legend.position="none") +
  scale_fill_brewer(palette = "Dark2")


caves_aspect <- cave %>% 
  mutate(TYPE = fct_relevel(TYPE, 
            "MP Caves", "All Caves")) %>%
  ggplot(aes(x = TYPE, y = areaASPECT_cut, fill = TYPE)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter() +
  ylab("degrees") + xlab("") +
  theme_cowplot() +
  theme(legend.position="none") +
  scale_fill_brewer(palette = "Dark2")

caves_distance <- cave %>% 
  mutate(TYPE = fct_relevel(TYPE, 
            "MP Caves", "All Caves")) %>%
  ggplot(aes(x = TYPE, y = areaDIST_cut2, fill = TYPE)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter() +
  ylab("meters") + xlab("") +
  theme_cowplot() +
  theme(legend.position="none") +
  scale_fill_brewer(palette = "Dark2")

plot_grid(caves_elevation, caves_slope, caves_aspect, caves_distance, labels=c("A", "B", "C", "D"), ncol = 2, nrow = 2)



```

```{r caves-statistics, eval = FALSE, echo = FALSE}

caves_elevation_MWU <- wilcox.test(areaDEM_cut ~ TYPE, data = cave,
                   exact = FALSE)

caves_slope_MWU <- wilcox.test(areaSLOPE_cut ~ TYPE, data = cave,
                   exact = FALSE)


caves_aspect_MWU <- wilcox.test(areaASPECT_cut ~ TYPE, data = cave,
                   exact = FALSE)


caves_distance_MWU <- wilcox.test(areaDIST_cut2 ~ TYPE, data = cave,
                   exact = FALSE)

variables <- c("Elevation", "Slope", "Altitude", "Distance")

u_statistics <- c(caves_elevation_MWU$statistic, caves_slope_MWU$statistic,
                   caves_aspect_MWU$statistic, caves_distance_MWU$statistic)

p_values <- c(caves_elevation_MWU$p.value, caves_slope_MWU$p.value,
                   caves_aspect_MWU$p.value, caves_distance_MWU$p.value)

caves_stats <- data.frame(variables, u_statistics, p_values)

caves_stats <- caves_stats %>% 
  rename(Variables = variables, `U statistic` = u_statistics, "p-value" = p_values) %>% 
  mutate(across(where(is.numeric), round, 2))

knitr::kable(caves_stats, caption = "Results of the Mann-Whitney U test, for each spatial variable, between caves with Middle Paleolithic occupations and all other archaeological caves identified in westernmost Iberia.")


```



## Principal Component Analysis

```{r PCA, eval = FALSE, echo = FALSE}


pca_data <- as.data.frame(sites_vals)

pca_data <- pca_data %>% 
  dplyr::select(SITE, Type, Dmn_R_M, Elevation = areaDEM_cut, Slope = areaSLOPE_cut, Aspect = areaASPECT_cut,
                Distance = areaDIST_cut2)

pca_data_active <- dplyr::select(pca_data, Elevation,
                                 Aspect, Slope, Distance)


# Test assumptions
pca_Cor <- cor(pca_data_active)

# Sampling adequacy or an appropriate number of observations relative to the number of variables being examined
assumptions <- paf(as.matrix(pca_data_active), eigcrit = 1, convcrit = .001)
round(print(assumptions$KMO), 3)


# Positive determinant of the correlation or variance-covariance matrices
round(det(pca_Cor), 3)


# Sphericity or the existence of the identity matrix
bartlettTest <- cortest.bartlett(pca_Cor, n = 54)
round(bartlettTest$p.value, 3)



# Plot Results

res.pca <- prcomp(pca_data_active, scale = TRUE)

groups <- as.factor(pca_data$Dmn_R_M)

fviz_pca_biplot(res.pca,
                col.ind = groups, # color by groups
                addEllipses = TRUE, # Concentration ellipses
                ellipse.type = "confidence",
                legend.title = "Raw Material",
                repel = TRUE,
                label = "var",
                pointsize = 3
) +
  theme(text = element_text(size=15), plot.title = element_text(lineheight=.8, face="bold"), plot.margin=unit(c(1,1,1,1),"cm"))



loadings_table <- round(as.data.frame(res.pca$rotation[,1:2]), 3)





```


# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
##### pagebreak

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

##### pagebreak

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
git2r::repository("../..")
```
